---
title: "Tutorial 4: Data Transformation and Analysis"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r install packages, include=FALSE, eval=FALSE}
# In order to run this document you need to install a package
# This can be done using the command below
# Click the play button at the top right of this box to run this 'chunk' of code

install.packages("nycflights13")

# Once installed, click the Run Document at the top of this window to begin the tutorial
```

```{r setup, include=FALSE}
library(nycflights13)
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
pulseData <- read.table("https://raw.githubusercontent.com/kellya72/R-tutorials/master/Tutorial%204/pulse.txt", header = TRUE)
fastFoodData <- read.csv("https://raw.githubusercontent.com/kellya72/R-tutorials/master/Tutorial%204/fastfood_calories.csv", header = TRUE)
playerData<- read.csv("https://raw.githubusercontent.com/kellya72/R-tutorials/master/Tutorial%205/player_data.csv", header = TRUE)
missingData <- pulseData[!complete.cases(pulseData),]
cleanPulseData <- na.omit(pulseData)
playerData <- mutate(playerData, years_active= year_end - year_start)
```


## Introduction


This weeks tutorial will show you how to read in datasets to R, save datasets as files and deal with missing data. We will also look at the `filter()`, `arrange()`,`select()`,`mutate()`, and `groupBy()` functions which are useful for data transformation.



### Loading Datasets


-   The function used to read in the individual files depends on the file type.
    -   `read.table()` can be used for reading in most file types which contain data in a table format. The function takes the form `read.table(file, header= FALSE, sep= "")` where `file` is the name of the file. The default value for `header` is `FALSE`, which means that the function assumes the data read in does not have a heading. If this is not the case simply set `header=TRUE`. The `sep` parameter assumes that the data is seperated in the file by a space, tab or new line. However, if the data is seperated by a `;` for example, set `sep= ";"`.
    
    -   `read.csv()` is used for reading in csv files. The function takes the form `read.csv(file, header= TRUE, sep=",")`. Note that `read.csv()` assumes the data has a heading, however you can change this by setting `header= FALSE`. Also it assumes that the data is seperated by a `,` but this parameter can also be changed.

- Run the following code to read in the ['pulse.txt'](https://raw.githubusercontent.com/kellya72/R-tutorials/master/Tutorial%204/pulse.txt) and ['fastfood\_calories.csv'](https://raw.githubusercontent.com/kellya72/R-tutorials/master/Tutorial%204/pulse.txt) files. 

- You will notice that there is a url in place of a file name in the following code. R can read this url and save the dataset, assigning it the variable name given. 

-   When reading in data, it is useful to assign names to the input as then the datasets can easily be included in other functions.

    
```{r readtable, exercise= TRUE, exercise.eval= FALSE}
pulseData <- read.table("https://raw.githubusercontent.com/kellya72/R-tutorials/master/Tutorial%204/pulse.txt", header = TRUE)
pulseData 
```



```{r readcsv, exercise= TRUE, exercise.eval= FALSE}
fastFoodData <- read.csv("https://raw.githubusercontent.com/kellya72/R-tutorials/master/Tutorial%204/fastfood_calories.csv", header = TRUE)
fastFoodData
```

- Alternatively files can be read in to R by saving them to your working directory and including the file name e.g. "pulse.txt" in the function. If the file was saved in another location and not the working directory, the full filepath would need to be included e.g. `read.table("user/downloads/pulse.txt")`. You must use forward slashes in the file path, R will return an error if backwards slashes are used.


### Saving Datasets

-   It is also possible to save datasets created or worked on in R as files.

-   The `read.table()` function can be used to do this. The function takes the form `read.table(x, file, sep = " ")` where `x` is the name of the dataset in R and `file` is name of the file you wish to create. The `sep` parameter once again is the method by which each row of data is seperated in the resulting file, with the default seperator set to a space.

-   For example to save the `pulseData` dataset as a file called `savedPulse.txt` you would do so using the following code:

`write.table(pulseData, "savedPulse.txt")`

- The file will be saved to the working directory you set during the R session. 

- This function is very useful if you have read an existing file into R, made changes to the dataset and then wish to save these changes by creating a new file containing the dataset. 






## Missing Data


-   Datasets often can contain missing data.

-   The `complete.cases()` function is used to identify the rows of data which are complete i.e. do not contain missing data.

-   Sometimes it is helpful to take a closer look at cases with missing data as opposed to just deleting them.

-   Run the following code to find any missing data in the `pulseData` dataframe and save them in a new dataframe called `missingData`.


```{r ex-pulse-complete, exercise= TRUE, exercise.eval= FALSE, }
missingData <- pulseData[!complete.cases(pulseData),]
missingData
```


-   The `!` in R is a negative operator, i.e. the code above is searching for cases that are **not** complete.

-   The method of accessing the missing rows of data is the same principle that was used previously when selecting rows in matrices (`matrix[rowsOfInterest, columnsOfInterest]`).

### Analysing the Missing Data

-   The `summary()` function is useful for quickly analysing a dataset.

```{r ex-pulse-summary, exercise= TRUE, exercise.eval= FALSE, }
summary(missingData)
```

-   In RStudio, to access an entire dataset, use the `View()` function.


### Removing Missing Data

-   After analysing the cases which contain missing elements, sometimes we then wish to remove them from our original dataset.

**Exercise: Use the `complete.cases()` function to create a new dataframe called `cleanPulseData` which contains no missing data.**

```{r ex-clean-data, exercise= TRUE, exercise.eval= FALSE}

```
```{r ex-clean-data-solution}
cleanPulseData <- pulseData[complete.cases(pulseData),]
```
-   An alternative method of removing missing data is to use the `na.omit()` function:

```{r ex-pulse-naomit, exercise= TRUE, exercise.eval= FALSE, }
cleanPulseData <- na.omit(pulseData)
```






## Data Transformation

-   The `filter()`, `arrange()` and `select()` functions are from the `dplyr` package which is a member of the `tidyverse` packages.

### `filter()`

-   The `filter()` function allows you to easily subset observations based on their values.

-   The data to be subsetted is first specified and the subsequent arguments are the expressions that filter the data frame.

-   For example, you may wish to create a subset of cases from `cleanPulseData` containing only individuals who smoke and weigh over 160.

```{r pulse-filter, exercise= TRUE, exercise.eval= FALSE}
smokersOver160 <- filter(cleanPulseData, 
                         RestingPulse == "Low", 
                         Smokes== "Yes", 
                         Weight> 160)
head(smokersOver160)
```


**Exercise: Create a subset from `cleanPulseData` which contains individuals with a low resting pulse who do not smoke and weight less than 180. Name the subset `lowRateNonSmokers`.**
```{r ex-pulse-filter, exercise= TRUE, exercise.eval= FALSE}

```
```{r ex-pulse-filter-solution}
lowRateNonSmokers <- filter(cleanPulseData, 
                            Smokes == "No", 
                            Weight < 180)
head(lowRateNonSmokers)
```



**Exercise: Create a subset containing individuals with a weight that is less than or equal to 170.**
```{r ex-pulse-subset, exercise= TRUE, exercise.eval= FALSE, }

```
```{r ex-pulse-subset-solution}
lessThanOrEqualTo170 <- filter(cleanPulseData, Weight <= 170)
head(lessThanOrEqualTo170)
```


### `arrange()`

-   The `arrange()` function sorts and orders the contents of a dataframe.

```{r pulse-arrange1, exercise= TRUE, exercise.eval= FALSE}
weightOrder <- arrange(cleanPulseData, Weight)
weightOrder[1:5,]
```


-   The data frame is arranged in ascending order by default. However, you can sort the data by descending order using the following code:

```{r pulse-arrange2, exercise= TRUE, exercise.eval= FALSE}
weightOrderDesc <- arrange(cleanPulseData, desc(Weight))
weightOrderDesc[1:5,]
```

-   It is also possible to include more than one column name in the `arrange()` function.

```{r pulse-arrange3, exercise= TRUE, exercise.eval= FALSE}
smokesAndWeightOrder <- arrange(cleanPulseData, Smokes, Weight)
smokesAndWeightOrder[1:10,]
```


**Exercise: Run the above function again but this time input `Weight` into the function before `Smokes`. What effect does this have on the resulting dataset?**
```{r ex-pulse-arrange, exercise= TRUE, exercise.eval= FALSE}

```
```{r ex-pulse-arrange-solution}
weightAndSmokesOrder <- arrange(cleanPulseData, Weight, Smokes)
weightAndSmokesOrder[1:10,]
```


### `select()`

-   The `select()` function allows you select only the variables you are interested in from a data frame.

-   For example, some datasets may contain hundreds of variables but you may only wish to analyse a few of them.

-   `fastFoodData` has 18 variables for each of its 515 observations. The code below shows how the `select()` function can be used to create a new dataset with less variables.

```{r pulse-select, exercise= TRUE, exercise.eval= FALSE}
fastFoodDataSimplified <- select(fastFoodData, 
                                 restaurant, 
                                 item, 
                                 calories)
head(fastFoodDataSimplified)
```

**Exercise: Use the `select()` function to create a data frame called `fastFoodNutrition` which contains the variables `restaurant`, `item`, `calories`, `total_fat`, `sugar` and `protein`.**
```{r ex-pulse-select, exercise= TRUE, exercise.eval= FALSE}

```
```{r ex-pulse-select-solution}
fastFoodNutrition <- select(fastFoodData, 
                            restaurant, 
                            item, 
                            calories, 
                            total_fat, 
                            sugar, 
                            protein)
head(fastFoodNutrition)
```





## More Transformations

- In this section we will be using another dataset called `playerData`.

```{r playerData, exercise= TRUE, exercise.eval= FALSE}
playerData<- read.csv("https://raw.githubusercontent.com/kellya72/R-tutorials/master/Tutorial%205/player_data.csv", header = TRUE)
head(playerData)
```


### `mutate()`

-   The `mutate()` function allows you create new columns (variables) that are functions of existing columns and adds them to the dataframe.

-   For example, the `playerData` dataset has two variables `year_start` and `year_end` which represent the year a player started their career and the year they stopped playing professionally. It is therefore possible to add a new column `years_active` to the existing dataset by doing the following:

``` {r years-active, exercise= TRUE, exercise.eval= FALSE}
playerData <- mutate(playerData, years_active= year_end - year_start)
head(playerData)
```

-   It is possible to create multiple new variables within the same `mutate()` function using the following format: `mutate(data, newVariable1, newVariable2, newVariable3, ...)`.

### `transmute()`

-   If you only wish to keep the new variables you have created, you can do so using the `transmute()` function as shown below:

``` {r years-active-transmute, exercise= TRUE, exercise.eval= FALSE}
transmute(playerData, years_active= year_end - year_start)
```

**Exercise: The `nycflights13` package contains the `flights` dataset.Add a new variable to the dataset called `kmPerMinute` by dividing the `distance` variable by the `air_time` variable.**

```{r kmPerMin, exercise= TRUE, exercise.eval= FALSE}

```

```{r kmPerMin-solution}
flights <- mutate(flights, kmPerMinute= distance/air_time)
head(flights)
```

### `group_by()`

-   The `group_by()` function groups entries in a dataset by given variables.

-   This is particularly useful when used in conjunction with the `summarise()` function.

-   Try running the following code which groups the players in the dataset by their college and then finds the average number of years players from different colleges are active.

``` {r group-by, exercise= TRUE, exercise.eval= FALSE}
byCollege <- group_by(playerData, college)
summarise(byCollege, averageYearsActive= mean(years_active))
```

**Exercise: Group the dataset using the `year_start` variable and then find the maximum `year_end` associated with each starting year.**

```{r group-by-ex1, exercise= TRUE, exercise.eval= FALSE}

```
```{r group-by-ex1-solution}
byStartYear <- group_by(playerData, year_start)
summarise(byStartYear, maxYearEnd= max(year_end))
```


**Exercise: Group the `flights` dataset by `dest` and `carrier` then find the average distance for each grouping.**

```{r group-by-ex2, exercise= TRUE, exercise.eval= FALSE}

```

```{r group-by-ex2-solution}
byDestAndCarrier <- group_by(flights, dest, carrier)
summarise(byDestAndCarrier, averageDistance= mean(distance))
```




Data Visualisation
------------------

-   Before plotting a graph it is often useful to employ some data manipulation techniques on a dataframe.

-   This allows us to create plots that are more specific which can aid us in data analysis.

-   This section should help to consolidate what you have already learned while also incorporating the new techniques from this week.

### Example

Look at the following code. Do you understand what the functions are doing and what the resulting graph is representing?

```{r graph-example, exercise= TRUE, exercise.eval= FALSE}
playersAfter1990 <- filter(playerData, year_start>1990)
playersAfter1990 <- group_by(playersAfter1990, year_start)
averageYearsActive <- summarise(playersAfter1990, meanYearsActive = mean(years_active))
ggplot(data = averageYearsActive) + 
geom_point(mapping = aes(x = year_start, y = meanYearsActive))
```



**Exercise: Using `playerData` create a boxplot comparing a players position and their weight. Note: In some cases players have switched positions and therefore their position values are equal to `G-F`, `F-C` etc. Do not alter the values, simply consider `G-F` as a seperate group to `G` and `F`.**

```{r graph-ex1, exercise= TRUE, exercise.eval= FALSE, warning=FALSE}

```
```{r graph-ex1-solution, warning=FALSE}
ggplot(data = playerData, mapping = aes(x= position, y = weight)) + 
  geom_boxplot()
```


**Exercise: Create a bar plot using the positions variable but only for players of height of 6-8. Colour the bars based on the position.**

```{r graph-ex2, exercise= TRUE, exercise.eval= FALSE}

```
```{r graph-ex2-solution}
players6ft8 <- filter(playerData, height == "6-8")
ggplot(data = players6ft8) + 
  geom_bar(mapping = aes(x = position, fill = position))
```

**Exercise: Group `playerData` by `college` and find the minimum `year_start` for each college. Create a bar plot of the number of colleges for each minimum start year up to and including 1955.**

```{r graph-ex3, exercise= TRUE, exercise.eval= FALSE}


```
```{r graph-ex3-solution}
groupByCollege <- group_by(playerData, college)
minYearStart <- summarise(groupByCollege, minYear = min(year_start))
minYearStart <- filter(minYearStart, minYear<=1955)
ggplot(data = minYearStart) + 
  geom_bar(mapping = aes(x = minYear))
```



For more information and examples on the functions used in this weeks tutorial and how to incorporate them in graphs, read the [data transformation](https://r4ds.had.co.nz/transform.html) and the [exploratory data analysis](https://r4ds.had.co.nz/exploratory-data-analysis.html) chapters from the [R for Data Science](http://r4ds.had.co.nz/index.html) book.

