---
title: "Tutorial 9: Factors"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This weeks tutorial will focus on factors. Factors are used for categorical variables. Categorical variables are variables which can take one of a limited and usually fixed number of possible outcomes. The colour of a car, blood group and breed of a dog are all examples of categorical variables. Sometimes variables are recoded into categories of categorical variables for data analysis purposes. For example, a dataset of houses might have a variable representing its price. The price variable can be altered so it instead labels the house as either 'cheap', 'average' or 'expensive' dependent on a specified range i.e all houses under 200,000 euro are labelled as cheap. The price variable is then considered to be a categorical variable. 

## The `forcats` Package
- The `forcats` package contains many useful functions for dealing with factors in R.

- You can read about all the functions contained in the package and their uses [here](https://cran.r-project.org/web/packages/forcats/forcats.pdf).

- Install and load the package in RStudio.

- Ensure you also load the `tidyverse` package. 

```{r, echo=FALSE, message=FALSE}
library(forcats)
library(tidyverse)
```


## Introduction to Factors
- Some categorical variables such as the days of the week or months have a rank order. 

- As you can see below, if these variables are not defined as factors then they are sorted by R alphabetically as opposed to their correct natural order. 

```{r}
days <- c("Monday", "Thursday", "Sunday", "Friday", "Saturday")
sort(days)
```
- This issue can be rectified by formatting the values as factors:

```{r}
daysLevels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
daysAsFactors <- factor(days, levels = daysLevels)
sort(daysAsFactors)
```
- The `daysLevels` vector represents all the possible values for the `days` variable and the correct order for the variable.

- Factors are also useful as they prevent the input of incorrect values for the variable and human error:

```{r}
days2 <- c("Tuesday", "Moday", "Saturday", "Sunday", "July")
days2AsFactors <- factor(days2, levels= daysLevels)
days2AsFactors
```
- As you can see the incorrect values are relabelled `NA`. You can find out more information about the incorrect values using the `parse_factor()` function. 

```{r}
days2AsFactors <- parse_factor(days2, levels = daysLevels)
```

- If given a dataset containing factors you can return the levels of the variable as follows:

```{r}
levels(days2AsFactors)
```

- This is useful as you may be dealing with datasets with thousands of observations. The `levels()` function allows you to quickly find out all the possible values of a given categorical variable. 

## The `alcohol_consumption` Dataset
- Download and read in alcohol_consumption.csv found in the **Tutorial 9** folder.

- Alternatively you can read in the dataset directly using the dataset url as shown below.

```{r}
alcoholData <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-06-26/week13_alcohol_global.csv")
```

## `fct_reorder()`
- `alcoholData` contains details of alcohol consumption in 195 different countries. 

- We want to display the `total_litres_of_pure_alcohol` variable in a scatterplot. 

- As the dataset is so large for our example we will only plot the first 10 countries.

```{r alcoholScatter}
subsetAlcohol <- alcoholData[1:10,]
ggplot(subsetAlcohol, aes(total_litres_of_pure_alcohol, country)) + geom_point()

```


- This plot would be easier to interpret if the countries were sorted based on the `total_litres_of_pure_alcohol` variable. This can be done using the `fct_reorder()` function. 

```{r alcoholScatterOrder}
ggplot(subsetAlcohol, aes(total_litres_of_pure_alcohol, fct_reorder(country,total_litres_of_pure_alcohol ))) + geom_point()

```

**Exercise 1: Randomly select 10 countries from the `alcoholData` dataset. Produce a scatterplot of with the `country` variable on y-axis and `spirit_servings` on the x-axis, ensuring that the countries are sorted based on their `spirit_servings` value. Hint: Use the `sample()` function to randomly select the countries.**

## Recoding Factors
### `fct_recode()`
- The `fct_recode()` function allows you to change the values of the levels and even to reduce the number of levels by renaming multiple levels to the same name.  

- The `thanksgivingMeal` dataset contains details about what different families eat for their Thanksgiving meal.

```{r}
thanksgivingMeal <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-11-20/thanksgiving_meals.csv")
levels(thanksgivingMeal$main_dish)
```

- By using the `levels()` function you can see all the different values for the `main_dish` variable. 

- You may want to adust the values for the `main_dish` variable.

```{r}
thanksgivingMeal$main_dish <- fct_recode(thanksgivingMeal$main_dish, "Beef"="Roast beef", "Other (please specify)"= "Turducken")
levels(thanksgivingMeal$main_dish)
```

**Exercise 2: Take a look at the different values for the `main_dish_other` variable. Use the `fct_recode()` function to tidy up some of the values so that the data will be cleaner and therefore subsequent analysis will be easier.**

### `fct_collapse()`
- Sometimes you may want to rename a number of values to just one single value in order to simplify data. 

- While this can be done using the `fct_recode()` function it is often quicker to use `fct_collapse()`. 

```{r}
levels(thanksgivingMeal$kids_table_age)
```

```{r}
thanksgivingMeal$kids_table_age <- fct_collapse(thanksgivingMeal$kids_table_age, "11 to 15" = c("11", "12", "13", "14", "15"), "16 to 20"= c("16", "17", "18", "19", "20") )
levels(thanksgivingMeal$kids_table_age)
```

**Exercise 3: Use the `fct_collapse()` function to reduce the number of different values in the `main_prep_other` variable.**

## The `summary()` Function
- The `summary()` function is a quick and easy way to gain an overall snapshot of a large dataset.

- You can input the whole dataset or just a single variable of interest into the function.

```{r}
summary(thanksgivingMeal)
summary(thanksgivingMeal$family_income)
```

- The `thanksgivingMeal` dataset only contains categorical data. The `summary()` function provides a lot of useful statistical data for numerical data also.

```{r}
summary(mpg$hwy)
```

